Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/Formalization.tex
Fiexed incoherence in declarations
Section 3

\section{Interaction of Interface Methods with Interface Composition}
From Java 8 interfaces can have three type of methods: abstract, default, and
static methods.
\begin{itemize}

\item Static methods are handled in a very clean way: they are visible only in
  the interface they are explicitly defined. This means the following code
  is ill-typed.
\begin{lstlisting}
interface A0 { static int m(){return 1;} }
interface B0 extends A0 {}
... B0.m()//ill typed
\end{lstlisting}
This is different from the way static methods are handled in classes. Here
static methods have simply no interaction with interface
composition (\Q@extends@ or \Q@implements@).

\item Abstract methods composition is accepted when exist a most specific one.
  For example, here method \texttt{Integer m()} from \texttt{B1} is visible in \Q@C1@.
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}

\item Default methods conflict with any other default or abstract method. For
  example the following code is rejected due to methods conflict.
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is different from what happens in most trait models, where \Q@D2@
would be accepted, and the implementation in \Q@A2@ would be part of the
behaviour of \Q@D2@.

\item A method in the current interface wins over any method in its
super-interfaces, provided that the method%  conforms to the subtype of all methods
% in its super-interfaces, i.e., the method
is the most specific one. This method also overrides conflict due to
inheritance. For example, the following code is accepted, but would be rejected
(see before) if the method \Q@m@ was not redefined in \texttt{D3} and
\texttt{E3}.
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { default int m(){return 42;} } //accepted
\end{lstlisting}

\end{itemize}

While trying to formally encode the Java specification we have done some tests to clarify corner case behaviour.
Consider the following correct declarations:
\begin{lstlisting}
interface A1{T m(); }
interface A2 extends A1{default T m(){ ... } }
interface A3 extends A2{T m(); }

@@

What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_i$? The 9 cases are shown here:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                        & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract, accepted                        & conservative error       & both abstract, accepted       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
We try to classify the results out of the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from \Q@A@${}_i$ and
  \Q@B@${}_i$ are both implemented, which is also an error in most trait models.
\item \textbf{both abstract, accepted} happens when the methods from \Q@A@${}_i$
  and \Q@B@${}_i$ are both abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method from \Q@A@${}_i$
  or \Q@B@${}_i$ is implemented, which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

During our experimentation, we found a bug in ECJ (Eclipse compiler for Java):
the case \textbf{M} extending \textbf{B3} and \textbf{A1} is accepted by
ECJ4.5.1 and rejected by javac.  By email communication with Brian Goetz
(leading Java 8 designer) we have confirmed that the expected behaviour is
rejection, hence this is a bug in ECJ.\footnote{This bug is fixed in the ECJ
  developer brunch , but not released as a stable version yet.}
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

@@

\noindent$\begin{array}{l}
\InTextAssert{35ex}{
\method\in\tops(\m,\Cs)}{
}\\
\InTextWith{
  \mBody(\m,\C)=\method,\ \method\mbox{ not a static method, and }}\\
\tab\tab  \C\in\Cs\ \mbox{ such that }
\forall \C'\in\Cs\setminus\C \mbox{ not } \C' \subtype \C

>>>>

\section{Interaction of Interface Methods with Interface Composition}
Before formalizing Classless Java, it is helpful to informally discuss the behaviour
of methods in Java 8 interfaces.

%We show several interesting cases
%and summarize compilation result of method composition into 3 categories:
%conflict error, accepted (both abstract) and conservative error. Meanwhile we
%also show different composition handling mechanism among traits, javac and ECJ.

\subsection{Methods in Java 8 Interfaces}
In Java 8 interfaces there are three types of methods: abstract, default, and
static methods.

\paragraph{Static methods} are handled in a very clean way: they are visible only in
  the interface in which they are explicitly defined. This means the following code
  is ill-typed.
\begin{lstlisting}
interface A0 { static int m(){return 1;} }
interface B0 extends A0 {}
... B0.m()//ill typed
\end{lstlisting}
This is different from the way static methods are handled in classes. Here
static methods have simply no interaction with interface
composition (\Q@extends@ or \Q@implements@).

\paragraph{Abstract method} composition is accepted when there exist a most specific one.
  For example, here method \texttt{Integer m()} from \texttt{B1} is visible in \Q@C1@.
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}

\paragraph{Default methods} conflict with any other default or abstract method. For
  example the following code is rejected due to method conflicts.
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is different from what happens in most trait models, where \Q@D2@
would be accepted, and the implementation in \Q@A2@ would be part of the
behaviour of \Q@D2@.

\paragraph{Resolving conflicts:}
A method in the current interface wins over any method in its
super-interfaces, provided that the method
is the most specific one. This method also overrides conflict due to
inheritance. For example, the following code is accepted, but would be rejected
(see before) if the method \Q@m@ was not redefined in \texttt{D3} and
\texttt{E3}.
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { default int m(){return 42;} } //accepted
\end{lstlisting}

\subsection{Classifying Method Composition}
When interfaces are composed and methods with the same name (and
signature) exist, there are 3 possible outcomes.
We illustrate the possible outcomes using the following (correct) declarations:
\begin{lstlisting}
interface A1{T m(); }
interface A2 extends A1{default T m(){ ... } }
interface A3 extends A2{T m(); }

@@

\noindent 
What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_j$? Nine representative cases are shown next:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                        & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract, accepted                        & conservative error       & both abstract, accepted       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
We try to classify the results in the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from are both implemented, which is also an error in most trait models.
\item \textbf{both abstract, accepted} happens when the methods from are both abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method from is implemented, which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

\paragraph{A bug:} During our experimentation, we found a bug in ECJ (Eclipse compiler for Java):
the case \textbf{M} extending \textbf{B3} and \textbf{A1} is accepted by
ECJ4.5.1 and rejected by javac.  By email communication with Brian Goetz
(leading Java 8 designer) we have confirmed that the expected behaviour is
rejection, hence this is a bug in ECJ. This bug was also reported by
  others and is fixed in the ECJ developer branch, but not released as a stable
  version yet.
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

@@

\noindent$\begin{array}{l}
\InTextDef{22ex}{
\method\in\tops(\m,\Cs)}{
}{
\mBody(\m,\C)=\method
}\\
\InTextWith{
 \method\mbox{ not a static method, and }
 \C\in\Cs\ \mbox{ such that }
\forall \C'\in\Cs\setminus\C \mbox{ not } \C' \subtype \C}
